\section{Conclusion}
This paper introduces a Causal Behavioral State Reasoning (CBSR) framework for APT detection in streaming provenance graphs. The core contribution is a formal 12-dimensional behavioral state space derived from online causal discovery on uncurated, potentially contaminated logs---eliminating the clean-data assumption that limits all existing provenance IDS methods. The Causal Intervention Score (CIS), grounded in graph boundary theory, provides a principled topological measure that separates localized administrative noise from multi-hop attack propagation. A multi-perspective parallel architecture (Local, Temporal, and Behavioral reasoning) analyzes this state space from complementary angles, while conformal prediction with online drift adaptation delivers distribution-free FPR guarantees. Across five diverse benchmarks, the framework achieves AUC of \textbf{0.9628} on real enterprise Windows logs (OpTC), AUC \textbf{1.0000} on Linux telemetry (BETH), and AUC \textbf{1.0000} on simulated datasets (TC3, NODLINK). On the challenging heterogeneous StreamSpot benchmark, it improves AUC from 0.4991 to \textbf{0.6909}---resolving the model-mismatch failure of pure causal approaches. At a 5\% FPR budget, conformal calibration controls empirical false alarm rates within 1.78\%--6.13\%, with SOC-practical top-$k$ alerting substantially improving operational precision on imbalanced datasets. Future work targets workload-conditioned SCM ensembles for heterogeneous graph streams, GPU-accelerated feature selection for very high-dimensional logs ($d > 1000$), and empirical evaluation of active SCM-aware mimicry attacks.
