% ============================================================================
% Abstract
% ============================================================================
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Modern provenance-based Intrusion Detection Systems (IDS) for detecting Advanced Persistent
Threats (APTs) share a critical deployment bottleneck: they require curated, pristine benign
training data to build normal behavioral profiles. In real-world enterprise environments, this
clean-data assumption fails due to stealthy attacker contamination, constant concept drift, and
prohibitive manual labeling costs.

This thesis presents \textbf{Causal Behavioral State Reasoning (CBSR)}, a provenance-based IDS
framework that infers latent cyber risk directly from uncurated, unlabeled streaming logs without
any clean-baseline collection period. CBSR projects raw system audit telemetry into a formal
12-dimensional \emph{Behavioral State Space} using online Structural Causal Models (SCMs)
discovered from potentially contaminated streams via RobustParCorr. Within this state space,
system dynamics are analyzed in parallel by three reasoning operators: (i) \emph{Local State
Reasoning} ($R_L$) that measures instantaneous causal mechanism violations; (ii) \emph{Temporal
State Reasoning} ($R_T$) that captures multi-stage attack progression via an LSTM sequence
model; and (iii) \emph{Behavioral State Reasoning} ($R_B$) that classifies complex attack
manifolds using a Behavioral Pattern Reasoner trained on sparse validation labels. Their outputs
are fused by a stacked meta-learner and calibrated with online split conformal prediction for
distribution-free false-alarm control under concept drift.

Novel contributions include: (1) a formal 12-dimensional causal Behavioral State Space grounded
in five invariant coordinate groups derived from online SCM outputs; (2) a Causal Intervention
Score (CIS) rooted in graph boundary theory that quantifies multi-hop attack propagation versus
localized administrative noise; and (3) a multi-perspective parallel reasoning architecture
that resolves the model-mismatch failures of pure causal detectors.

Evaluated on five diverse benchmarks, CBSR achieves AUC of \textbf{0.9628} on real enterprise
Windows logs (DARPA OpTC), AUC \textbf{0.9901} on Linux host telemetry (BETH), and AUC
\textbf{0.9837} and \textbf{0.9784} on simulated multi-stage APT campaigns (TC3, NODLINK). On the challenging
heterogeneous StreamSpot benchmark, the framework improves AUC from 0.4991 to \textbf{0.6909}.
Conformal calibration holds empirical false alarm rates within 1.78\%--6.13\% at a 5\% FPR
budget, and a 13$\times$ runtime speedup confirms streaming deployment readiness.

\vspace{0.5cm}
\noindent\textbf{Keywords:} Intrusion Detection, Advanced Persistent Threats, Causal Discovery,
Behavioral State Space, Conformal Prediction, Provenance Graphs, Structural Causal Models,
Semi-Supervised Detection
